Wigglyhedra

### 2024-06-07 draft/wigglyhedra.tex
Added comment; also typo fixes and readability.

\begin{theorem}
\label{thm:wigglyFan}
The collection of cones~$\R_{\ge 0} \langle \b{g}(A) \rangle$ for all~$A$ in the wiggly complex~$\wigglyComplex_n$ forms a complete simplicial fan of~$\HH_{2n}$, called \defn{wiggly fan}~$\wigglyFan_n$.
\end{theorem}

@@

\begin{remark}
In their work on polytopal realizations of $\b{g}$-vector fans of finite type cluster algebras~\cite{HohlwegPilaudStella}, C.~Hohlweg, V.~Pilaud and S.~Stella already constructed a \defn{universal associahedron}~$\polytope{U}_n$ which somehow contains all Cambrian fans (actually all $\b{g}$-vector fans from any initial cluster seed).
More precisely, the $\delta$-Cambrian fan for each~$\delta \in \{+,-\}^n$  appears as a section of the normal fan of the universal associahedron~$\polytope{U}_n$, which is a $(n^2+n-4)/2$-dimensional polytope.
In contrast, \cref{prop:CambrianFan} states that the $\delta$-Cambrian fan for each~$\delta \in \{+,-\}^n$ appears as a link of the normal fan of the wigglyhedron~$\wigglyhedron_n$, which is a $(2n-1)$-dimensional polytope.
There seem to be no connection between these two universal polytopes.
However, this certainly raises the question to construct analogues of the wigglyhedron for all finite Coxeter groups.
\end{remark}

@@

\begin{definition}
  The zigzag algebra \(A_m\) corresponding to the quiver \(Q_m\) depicted in~\cref{fig:am-quiver} is the quotient of \(kQ_m\) by the following relations:
  \begin{itemize}
  \item \((i|j|l) = 0\) whenever \(i \neq l\);
  \item \((i|j|i) = (i|l|i)\) for every \(j\) and \(l\) such that there are arrows between \(i\) and \(j\) and between \(i\) and \(l\).
  \end{itemize}
\end{definition}

\begin{remark}
  The definition of the zigzag algebra as used here originally appeared in~\cite[Section 3]{hue.kho:01}.
  This is an ``un-signed'' version of the zigzag algebra.
  It is technically better to use a ``signed'' version of the zigzag algebra, and the comparison between the two is discussed in~\cite[Remark 6.6]{bap.deo.lic:20}.
  As the two versions are isomorphic in our example, we ignore this technicality and use the easier definition.
\end{remark}

@@

Recall that an object of the form \(M_i[h]\) is placed in homological degree \(-h\);
since the differential \(d\) raises the homological degree by one, it \emph{lowers} the homological shift by one.
That is, we have
\[d \colon M_i\langle g \rangle[h] \to \bigoplus M_{i'}\langle g' \rangle[h-1].\]
In other words, \(d\) maps \(P^h\) to \(P^{h-1}\).
\begin{definition}\label{def:linear-complex}
  Let \(P^{\bullet}\) be an object of \(\mathcal{C}_m\), described as a bigraded module
  \[P^{\bullet} = \bigoplus M_i\langle g \rangle[h]\]
  together with an internal differential \(d \colon P^{\bullet} \to P^{\bullet + 1}\).
  We say that \(P^{\bullet}\) has \emph{level \(\ell\)} for some \(\ell \in \mathbb{Z}\) if for each bigraded piece \(M_i \langle g \rangle[h]\) of \(P^{\bullet}\), we have \(g + h = \ell\).
  We say that \(P^{\bullet}\) is \emph{linear} if \(P^{\bullet}\) has level \(\ell = 0\).

@@

Also recall the following statements for future use.
\begin{defprop}\label{defprop:cohomology}
  Let \((\mathcal{C}^{\le 0}, \mathcal{C}^{\geq 0})\) be a bounded \(t\)-structure on a triangulated category \(\mathcal{C}\) with heart \(\mathcal{A} = \mathcal{C}^{\leq 0} \cap \mathcal{C}^{\geq 0}\).
  For each integer \(i\), we have the \(i\)th truncation functors
  \[\tau^{\leq i} \colon \mathcal{C} \to \mathcal{C}^{\leq i}\]
  and
  \[\tau^{\geq i} \colon \mathcal{C} \to \mathcal{C}^{\geq i},\]
  which are left and right adjoint respectively\asilata{Check left vs right.} to the inclusion functors \(\mathcal{C}^{\leq 0} \to \mathcal{C}\) and \(\mathcal{C}^{\geq 0} \to \mathcal{C}\).
  The \(i\)th cohomology functor is defined to be the composition \(\tau^{\leq i} \circ \tau^{\geq i}\).
  It is equal to the composition \(\tau^{\geq i} \circ \tau^{\leq i}\), and maps an object of \(\mathcal{C}\) to an object of \(\mathcal{A}\).
\end{defprop}
In the presence of a bounded \(t\)-structure, every object of \(\mathcal{C}\) has a canonical filtration whose factors are shifts of its cohomology objects.
\begin{defprop}\label{defprop:cohomology-filtration}
  Let \(\mathcal{C}\) be any triangulated category, and \(\mathcal{A} \subset \mathcal{C}\) the heart of a bounded \(t\)-structure on \(\mathcal{C}\).
  Let \(X\) be a non-zero object of \(\mathcal{C}\), and let \(k\) and \(l\) be the maximum and minimum integers respectively for which \(H^k(X)\) and \(H^{l}(X)\) are non-zero.
  \asilata{Find references for cohomology and cohomology filtration.}
  Then \(X\) has a unique filtration of the form
  \[
    \begin{tikzcd}[column sep=0.1cm]
      0  \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k+1} \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} && \cdots \arrow[dashed]{ul} &&H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]  
\end{defprop}

@@

Recall from~\cref{defprop:cohomology-filtration} that any object \(X\) of \(\mathcal{C}_m\) has a canonical cohomology filtration whose factors are homological shifts of objects in \(\mathcal{A}_m\).
For general \(X\), the cohomology filtration can be arbitrary.
However, it turns out that the cohomology filtration is highly constrained if \(X\) is spherical, and these constraints are given by wiggly pseudotriangulations, as explained in the following proposition.
\begin{proposition}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} & \cdots & H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of all nonzero objects \(\{H^j(X)\}\) taken without multiplicity, and let \(T\) be corresponding set of wiggly curves.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of this proposition is an easy consequence of analysing cohomology filtration from the point of view of arcs on the disk with marked points.
  \asilata{TODO.}
\end{proof}

\subsection{Categorical constraints on compatibility of curves}
In this section, we give a categorical criterion for two wiggly arcs to be compatible as in~\cref{def:compatible}.
As such, this section is somewhat dual to the previous section --- our criterion determines whether a pair of wiggly arcs can belong together in a single wiggly pseudotriangulation, whereas the previous section generated a wiggly pseudotriangulation given any object of the category.
\begin{proposition}
  Consider two wiggly arcs \(\alpha\) and \(\alpha'\), and let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to \(\alpha\) and \(\alpha'\) respectively.
  Then \(\alpha\) and \(\alpha'\) are compatible in the sense of~\cref{def:compatible} if and only if \(\Ext^1(X\langle g \rangle, X\langle g' \rangle) = 0\) for any two integers \(g\) and \(g'\).
\end{proposition}
\begin{remark}
  This proposition is similar to the condition of being compatible in a cluster-tilting object.
  \asilata{Expand and make precise.}
\end{remark}
\begin{remark}
  As stated, this proposition is not expected to be true in greater generality.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
  \asilata{Decide what to say here, because the forthcoming work is far from complete, although I do have counterexamples in type \(D_4\). Discussing how to interpret these counterexamples may not involve too much more work?}
\end{remark}
\begin{question}
  How to best relate this section to the previous one purely categorically?
  That is, given the categorical constraint on compatibility of arcs, is it also clear that it gives a constraint on objects appearing in the cohomology filtration?
  \asilata{Make more precise.}
\end{question}

>>>>

\begin{theorem}
\label{thm:wigglyFan}
The collection of cones~$\R_{\ge 0} \langle \b{g}(A) \rangle$ for all~$A$ in the wiggly complex~$\wigglyComplex_n$ forms a complete simplicial fan of~$\HH_{2n}$, called the \defn{wiggly fan}~$\wigglyFan_n$.
\end{theorem}

@@

\begin{remark}
In their work on polytopal realizations of $\b{g}$-vector fans of finite type cluster algebras~\cite{HohlwegPilaudStella}, C.~Hohlweg, V.~Pilaud and S.~Stella already constructed a \defn{universal associahedron}~$\polytope{U}_n$ which somehow contains all Cambrian fans (actually all $\b{g}$-vector fans from any initial cluster seed).
More precisely, the $\delta$-Cambrian fan for each~$\delta \in \{+,-\}^n$  appears as a section of the normal fan of the universal associahedron~$\polytope{U}_n$, which is a $(n^2+n-4)/2$-dimensional polytope.
In contrast, \cref{prop:CambrianFan} states that the $\delta$-Cambrian fan for each~$\delta \in \{+,-\}^n$ appears as a link of the normal fan of the wigglyhedron~$\wigglyhedron_n$, which is a $(2n-1)$-dimensional polytope.
There seem to be no connection between these two universal polytopes.
However, this certainly raises the question to construct analogues of the wigglyhedron for all finite Coxeter groups.
\asilata{I can come up with a possible categorical pathway towards the wiggly flip graph for arbitrary finite Coxeter groups, related to~\cref{prop:hn-filtration}. But I have no idea how to construct the polytope!}
\end{remark}

@@

\begin{definition}
  The \defn{zigzag algebra} \(A_m\) corresponding to the quiver \(Q_m\) depicted in~\cref{fig:am-quiver} is the quotient of \(kQ_m\) by the following relations:
  \begin{itemize}
  \item \((i|j|l) = 0\) whenever \(i \neq l\);
  \item \((i|j|i) = (i|l|i)\) for every \(j\) and \(l\) such that there are arrows between \(i\) and \(j\) and between \(i\) and \(l\).
  \end{itemize}
\end{definition}

\begin{remark}
  The definition of the zigzag algebra as used here originally appeared in~\cite[Section 3]{hue.kho:01}.
  This is the ``un-signed'' zigzag algebra.
  It is technically better to use a ``signed'' zigzag algebra, and the comparison between the two is discussed in~\cite[Remark 6.6]{bap.deo.lic:20}.
  As the two versions are isomorphic in our example, we ignore this technicality and use the easier definition.
\end{remark}

@@

Recall that an object of the form \(M_i[h]\) is placed in homological degree \(-h\);
since the differential \(d\) raises the homological degree by one, it \emph{lowers} the homological shift by one.
That is, we have
\[d \colon M_i\langle g \rangle[h] \to \bigoplus M_{i'}\langle g' \rangle[h-1].\]
In other words, \(d\) maps \(P^h\) to \(P^{h-1}\).
\begin{definition}\label{def:linear-complex}
  Let \(P^{\bullet}\) be an object of \(\mathcal{C}_m\), described as a bigraded module
  \[P^{\bullet} = \bigoplus M_i\langle g \rangle[h]\]
  together with an internal differential \(d \colon P^{\bullet} \to P^{\bullet + 1}\).
  We say that \(P^{\bullet}\) has \defn{level \(\ell\)} for some \(\ell \in \mathbb{Z}\) if for each bigraded piece \(M_i \langle g \rangle[h]\) of \(P^{\bullet}\), we have \(g + h = \ell\).
  We say that \(P^{\bullet}\) is \defn{linear} if \(P^{\bullet}\) has level \(\ell = 0\).

@@

Also recall the following statements for future use.
\begin{defprop}\label{defprop:cohomology}
  Let \((\mathcal{C}^{\le 0}, \mathcal{C}^{\geq 0})\) be a bounded \(t\)-structure on a triangulated category \(\mathcal{C}\) with heart \(\mathcal{A} = \mathcal{C}^{\leq 0} \cap \mathcal{C}^{\geq 0}\).
  For each integer \(i\), we have the \(i\)th truncation functors
  \[\tau^{\leq i} \colon \mathcal{C} \to \mathcal{C}^{\leq i}\]
  and
  \[\tau^{\geq i} \colon \mathcal{C} \to \mathcal{C}^{\geq i},\]
  which are left and right adjoint respectively\asilata{Check left vs right.} to the inclusion functors \(\mathcal{C}^{\leq 0} \to \mathcal{C}\) and \(\mathcal{C}^{\geq 0} \to \mathcal{C}\).
  The \(i\)th \defn{cohomology functor} is defined to be the composition \(\tau^{\leq i} \circ \tau^{\geq i}\).
  It is equal to the composition \(\tau^{\geq i} \circ \tau^{\leq i}\), and maps an object of \(\mathcal{C}\) to an object of \(\mathcal{A}\).
\end{defprop}
In the presence of a bounded \(t\)-structure, every object of \(\mathcal{C}\) has a canonical filtration whose factors are shifts of its cohomology objects.
\begin{defprop}\label{defprop:cohomology-filtration}
  Let \(\mathcal{C}\) be any triangulated category, and \(\mathcal{A} \subset \mathcal{C}\) the heart of a bounded \(t\)-structure on \(\mathcal{C}\).
  Let \(X\) be a non-zero object of \(\mathcal{C}\), and let \(k\) and \(l\) be the maximum and minimum integers respectively for which \(H^k(X)\) and \(H^{l}(X)\) are non-zero.
  \asilata{Find references for cohomology and cohomology filtration.}
  Then \(X\) has a unique \defn{cohomology filtration}, which is the unique filtration of the form
  \[
    \begin{tikzcd}[column sep=0.1cm]
      0  \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k+1} \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} && \cdots \arrow[dashed]{ul} &&H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]  
\end{defprop}

@@

Recall from~\cref{defprop:cohomology-filtration} that any object \(X\) of \(\mathcal{C}_m\) has a canonical cohomology filtration whose factors are homological shifts of objects in \(\mathcal{A}_m\).
For general \(X\), the cohomology filtration can be arbitrary.
However, it turns out that the cohomology filtration is highly constrained if \(X\) is spherical, and these constraints are given by wiggly pseudotriangulations, as explained in the following proposition.
\begin{proposition}\label{prop:hn-filtration}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} & \cdots & H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of all nonzero objects \(\{H^j(X)\}\) taken without multiplicity, and let \(T\) be corresponding set of wiggly curves.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of this proposition is an easy consequence of analysing cohomology filtration from the point of view of arcs on the disk with marked points.
  \asilata{TODO.}
\end{proof}

\subsection{Categorical constraints on compatibility of curves}
In this section, we give a categorical criterion for two wiggly arcs to be compatible as in~\cref{def:compatible}.
As such, this section is somewhat dual to the previous section --- the criterion in this section determines whether a pair of wiggly arcs can belong together in a single wiggly pseudotriangulation, whereas the previous section generated a wiggly pseudotriangulation given any object of the category.
\begin{proposition}
  Consider two wiggly arcs \(\alpha\) and \(\alpha'\), and let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to \(\alpha\) and \(\alpha'\) respectively.
  Then \(\alpha\) and \(\alpha'\) are compatible in the sense of~\cref{def:compatible} if and only if \(\Ext^1(X\langle g \rangle, X\langle g' \rangle) = 0\) for any two integers \(g\) and \(g'\).
\end{proposition}
\begin{proof}
  \asilata{TODO.}
\end{proof}
\begin{remark}
  This proposition is similar to the condition of being compatible in a cluster-tilting object.
  \asilata{Expand and make precise.}
\end{remark}
\begin{remark}
  As stated, this proposition is not expected to be true in greater generality.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
  \asilata{Decide what to say here, because the forthcoming work is far from complete, although I do have counterexamples in type \(D_4\). Discussing how to interpret these counterexamples may not involve too much more work?}
\end{remark}
\begin{question}
  How to best relate this section to the previous one purely categorically?
  That is, given the categorical constraint on compatibility of arcs, is it also clear that it gives a constraint on objects appearing in the cohomology filtration?
  \asilata{Make more precise.}
\end{question}
